# Understanding population attack rates from population survey 

Introduction

```{r setup}
here()
test_duration = 8
```

Modelling Method

-   prevalence (P) of PCR-test-positivity at a given time-point is converted to daily incidence (I) at that point, assuming a typical duration of PCR positivity (T). : I = P/T (caveat: ignoring epidemic growth biases)

-   

Netherlands RIVM

-   Study details: needs translation

```{r}
### some inspiration
# https://x.com/adamjkucharski/status/1746913088559808744?s=20

# https://www.rivm.nl/coronavirus-covid-19/onderzoek/infectieradar
# specifically here: https://dashboard.infectieradar.nl/data/data_positive_tests.csv
# must translate properly
# must go through protocol

path_data="../../../Big input data/data_positive_tests.csv"
x=read.csv(path_data,sep = ";",dec = ",") %>% mutate(WEEK=ymd(WEEK))
x = x %>% select(date=WEEK,
             value=POS_TEST_PERC) %>% 
  mutate(location="NL")
df_ii = x
###########################
# date, value, location
df_nl = x

df_nl$date %>% diff() %>% table()
st = paste("Data set is properly weekly"); cat(green(st))

df_nl = df_nl %>% 
  mutate( average_value=rollmean(value,k=52,align = "right",fill=NA) ) %>% 
  mutate( average=rollsum(value,k=52,align = "right",fill=NA)/test_duration ) 
df_nl$average %>% mean(na.rm = T) # 8%
df_nl$average_value

df_nl %>% filter(date>"2022-10-01",date<"2024-02-03") %>% 
  ggplot(aes(date,value)) + geom_line() + coord_cartesian(ylim=c(0,NA))
df_nl %>% filter(date>"2022-10-01",date<"2024-02-03") %>% 
  ggplot(aes(date,average)) + geom_line() + coord_cartesian(ylim=c(0,NA))

```

England ONS\

-   Measuring: percent testing positive for SARS-CoV-2 (PCR-test-positivity prevalence)

-   Range is about **1 year**: 16 January 2022 to 31 January 2023

-   Eyeballing/expectation: around **3% prevalence**

-   Final dataset from March 2023, then dicontinued

-   Study details

    -   private housholds in England, Norther Ireland, Wales, Scotland

```{r}
# notice about discontinuation
# https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/bulletins/coronaviruscovid19infectionsurveypilot/latest

# on this webstie
# https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/bulletins/coronaviruscovid19infectionsurveypilot/10february2023#coronavirus-covid-19-infection-survey-data

# in the section: Estimated percentage of the population testing positive for COVID-19 on nose and throat swabs
# https://www.ons.gov.uk/visualisations/dvc2462/national/datadownload.xlsx

# this excel file is super messy 
# I combined England, Wales, Nothern Ireland, Scotland 
# modelled percent into a single sheet ("together")

path_data="../../../Big input data/datadownload_ons.xlsx"
x=read_xlsx(path_data,sheet = "together")
x %>% ggplot(aes(date,positive,col=region)) + geom_line() 
x %>% filter(region=="England") %>% pull(date) %>% range()

df_uk = x


df_uk = df_uk %>% 
  group_by(region) %>% 
  mutate( average=rollsum(positive,k=52,align = "right",fill=NA)/test_duration ) 
df_uk %>% filter(!is.na(average))

st = paste("Around 20 to 24% attack rate"); cat(green(st))

df_uk %>% filter(date>"2022-10-01",date<"2024-02-03") %>% 
  ggplot(aes(date,average,col=region)) + geom_line() + coord_cartesian(ylim=c(0,NA))

```

Germany - Uni Mainz

```{r}
# website: https://www.unimedizin-mainz.de/sentisurv/startseite/startseite.html

```
