# Understanding population attack rates from population surveys 

## Introduction

-   Getting a close to the true infection incidence/prevalence/attack rate is a crucial ingredient for scenario modelling

-   Population surveys that use randomised testing for a disease are the best data source to understand infection incidence/prevalence/attack rate.

```{r setup}
here()
test_duration = 8
```

## Modelling Method and data sources

-   Prevalence (P) of PCR-test-positivity at a given time-point is converted to daily incidence (I) at that point, assuming a typical duration of PCR positivity (T). : I = P/T (with caveats). The yearly infection attack rate (AR) is then given through AR = 365\*I

-   We searched the web for things like "countries random testing COVID-19"

    -   we identified 3 countries with ongoing or recent population testing for covid: UK, Germany, Netherlands

    -   we identified a review paper from Mid 2021: 37 surveys from 19 countries [[Population‐based prevalence surveys during the Covid‐19 pandemic: A systematic review - PMC (nih.gov)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7883186/)]

### Netherlands RIVM

-   Study details: needs translation

```{r}
### some inspiration
# https://x.com/adamjkucharski/status/1746913088559808744?s=20

# https://www.rivm.nl/coronavirus-covid-19/onderzoek/infectieradar
# specifically here: https://dashboard.infectieradar.nl/data/data_positive_tests.csv
# must translate properly
# must go through protocol

path_data="../../../Big input data/data_positive_tests.csv"
x=read.csv(path_data,sep = ";",dec = ",") %>% mutate(WEEK=ymd(WEEK))
x = x %>% select(date=WEEK,
             value=POS_TEST_PERC) %>% 
  mutate(location="NL")
df_ii = x
###########################
# date, value, location
df_nl = x

df_nl$date %>% diff() %>% table()
st = paste("Data set is properly weekly"); cat(green(st))

df_nl = df_nl %>% 
  mutate( average_value=rollmean(value,k=52,align = "right",fill=NA) ) 
prev_nl = df_nl$average_value %>% mean(na.rm = T) 

ar_nl = df_nl$average_value/8 * 365
ar_nl %>% quantile(probs=c(0.25,0.5,0.75),na.rm=T)

df_nl %>% filter(date>"2022-10-01",date<"2024-02-03") %>% 
  ggplot(aes(date,value)) + geom_line() + coord_cartesian(ylim=c(0,NA))
df_nl %>% filter(date>"2022-10-01",date<"2024-02-03") %>% 
  ggplot(aes(date,average_value)) + geom_line() + coord_cartesian(ylim=c(0,NA))

```

### England ONS 

-   Measuring: percent testing positive for SARS-CoV-2 (PCR-test-positivity prevalence)

-   Range is about **1 year**: 16 January 2022 to 31 January 2023

-   Eyeballing/expectation: around **3% prevalence**

-   Final dataset from March 2023, then dicontinued

-   Study details

    -   private housholds in England, Norther Ireland, Wales, Scotland

```{r}
# notice about discontinuation
# https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/bulletins/coronaviruscovid19infectionsurveypilot/latest

# on this webstie
# https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/bulletins/coronaviruscovid19infectionsurveypilot/10february2023#coronavirus-covid-19-infection-survey-data

# in the section: Estimated percentage of the population testing positive for COVID-19 on nose and throat swabs
# https://www.ons.gov.uk/visualisations/dvc2462/national/datadownload.xlsx

# this excel file is super messy 
# I combined England, Wales, Nothern Ireland, Scotland 
# modelled percent into a single sheet ("together")

path_data="../../../Big input data/datadownload_ons.xlsx"
x=read_xlsx(path_data,sheet = "together") %>% 
  mutate(date=ymd(date))
x %>% ggplot(aes(date,positive,col=region)) + geom_line() 
x %>% filter(region=="England") %>% pull(date) %>% range()

df_uk = x

prevs = list(); list_i = 1
location_v = unique(df_uk$region)
for (location in location_v){
  df_i = df_uk %>% filter_log(region==location)
  df_sorted = df_i %>% arrange(desc(date))
  this_works = T; i = 1
  while (this_works){
    recent_date = df_sorted$date[i]
    date_ayear_ago = recent_date - 365 
    closest_date_ayear_ago = df_sorted$date[which.min(abs(df_sorted$date - date_ayear_ago) )]
    day_diff = as.numeric(closest_date_ayear_ago - date_ayear_ago) %>% abs()
    
    prev_i = df_sorted %>% filter(date>=closest_date_ayear_ago & 
                           date<=recent_date) %>% 
      pull(positive) %>% mean()
    
    prevs[[list_i]] = 
      tibble(location,
           date_end=recent_date,
           date_start=closest_date_ayear_ago,
           prev=prev_i
           )
    list_i = list_i + 1
    i = i + 1
    if (day_diff>50) this_works = F
  }
}
df_prevs = bind_rows(prevs) 

df_prevs %>% mutate(id=1:n()) %>% 
  ggplot(aes(y=prev,yend=prev,x=date_start,xend=date_end)) +
  geom_segment(aes(col=location)) + facet_wrap(~location)

attack_rate = (df_prevs$prev)/8 * 365
attack_rate %>% quantile(probs=c(0.1,0.5,0.9)) %>% round(1)

df_prevs %>% mutate(ar=prev/8*365) %>% column_stats_ingroups(ar,location) %>% 
  pivot_wider(names_from = quant,values_from = val)

st = paste("Around 150% to 175% attack rate"); cat(green(st))


```

### Germany - Uni Mainz 

```{r}
# website: https://www.unimedizin-mainz.de/sentisurv/startseite/startseite.html
# not for download
# I emailed
# in the meantime eye-balling
incidence_7days = 1000/(100*1e3)
incidence_daily = incidence_7days/7
attack_rate_mainz = incidence_daily*365

```

```{r}
# all together
manual_vals = function(df,vals){
  df$lower = vals[1]
  df$point = vals[2]
  df$upper = vals[3]
  return(df)
}
set_loc_mode = function(df,vals,name) {
  df$mode = (vals[1])
  return(df)
}


tibble(mode=0,lower=0,point=0,upper=0) %>% 
  ggplot( aes(y=mode,xmin=lower,xmax=upper,x=point) ) + 
  #
  geom_pointrange(data=.%>%manual_vals( c(151.6, 153.2, 153.3) )%>%
                    set_loc_mode(0.0,"England - ONS"),col="darkblue" ) + # 
  geom_pointrange(data=.%>%manual_vals( c(151.9, 152.9, 153.7) )%>%
                    set_loc_mode(-.1,"Wales - ONS"),col="darkblue" ) + # 
  geom_pointrange(data=.%>%manual_vals( c(171.6, 173.1, 175.1) )%>%
                    set_loc_mode(-.2,"Northern Ireland - ONS"),col="darkblue" ) + #
  geom_pointrange(data=.%>%manual_vals( c(171.6, 173.1, 175.1) )%>%
                    set_loc_mode(-.3,"Scotland - ONS"),col="darkblue" ) + #
  geom_pointrange(data=.%>%manual_vals( c(52, 52, 52) )%>%
                    set_loc_mode(-.4,"Rhineland-Palatinate - UMainz"),col="darkgrey" ) + # 
  geom_pointrange(data=.%>%manual_vals( c(43.2, 56.3, 78.6) )%>%
                    set_loc_mode(-.5,"Netherlands - RIVM"),col="orange" ) + # 
  #
  geom_vline(xintercept = 100,linetype="dashed") +
  scale_y_continuous(breaks = c(0.0,-0.1,-0.2,-0.3,-0.4,-0.5),position = "right",
                     labels = c("England - ONS",
                                "Wales - ONS",
                                "Northern Ireland - ONS",
                                "Scotland - ONS",
                                "RP - UMainz",
                                "Netherlands - RIVM") ) +
  coord_cartesian(xlim=c(0,NA)) + 
  labs(caption = "Yearly attack rate estimates (with 50% bound)",y="",x="")

```
