---
title: "season_boe"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
editor_options: 
  chunk_output_type: console
  chunk_output_highlight: default
---

# *Season BOE: A Simple COVID-19 Back-of-the-Envelope Model*

This was later moved into the COVID19_modelling repo.

## *Introduction*

-   *Given that SARS-CoV2 immunity as reached 'equilibrium', can we model viral spread in a very simplified way?*

-   *high viral attack rate (likely above 50% per year) relative to typical speed of waning (substantial waning within 1/2 year)*

-   *immunity has reached an equilibrium: majority of individuals in EU/EEA have had multiple exposures and vaccines, and for impact of infection on transmission the immunity history does not matter*

-   *Thus, infection impact on population immunity can be modelled similarity for 22/23 as for 23/24*

*\*absence of NPIs, and less prevalence-elastic behaviour -\> Rt mostly impacted by immunity, seasonality, variants*

*\*by accounting for immunity (and variants) can we estimate seasonality?*

## *Method assumptions in very brief*

-   *Infection-intensity indicator is proportional to the true infection incidence (clearly NOT true with reported cases)*

-   *Viral transmission is only driven through seasonality, immunity accumulation and waning (and new variants)*

-   *Thus, at a similar time of the year, a similar Rt value implies a similar level of prevailing population-immunity (assuming no impact of variants)*

-   *For example we could say: in Nov 2022 and in Nov 2023 when Rt was close to 1, population-immunity was nearly the same*

-   *Immunity waning occurs in a relatively simple way ( 100%-immunity-compartment -\> X%-immunity-compartment -\> susceptible with exponential flow )*

-   *Parametric assumptions*

-   *Serial interval (si)=3.6 days*

-   *R0 (r0) = 6*

-   *Yearly infection attack-rate (ar): 60%*

-   *Reduction in susceptibility through infection (re): 90%*

## *Conceptual model walk through*

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(here)
# set R up 
source("../setup.R")
source("settings.R")
source("functions_specific.R")
source("checks.R")
```

```{r}
compute_Rt_without_immunity = function( par,
                                        # ability to fix parameters
                                        wane_imm_fract_weekly_1_log=NA,
                                        Imm_2_partial=NA,
                                        wane_imm_fract_weekly_2_log=NA,
                                        #
                                        df_Rt_value_pop,
                                        for_optim=F,
                                        p,
                                        days_to_week,
                                        sim_o=NA) {
  # df_Rt_value_pop must be daily
  # we apply a chunking-of-days technique to speed up optim etc
  simulate = !is.na(sim_o)[1] # logical to determine whether or not to forward simulate 
  ## recasting parameters (needed due to optim use)
  p$ar_and_re = exp(par[1])
  wane_imm_fract_weekly_1_logit = wane_imm_fract_weekly_1_log
  if (is.na(wane_imm_fract_weekly_1_log)) wane_imm_fract_weekly_1_logit = par[2]
  Imm_2_partial_logit = Imm_2_partial
  if (is.na(Imm_2_partial)) Imm_2_partial_logit = par[3]
  wane_imm_fract_weekly_2_logit = wane_imm_fract_weekly_2_log
  if (is.na(wane_imm_fract_weekly_2_log)) wane_imm_fract_weekly_2_logit = par[4]
  
  
  # waning compartments could be further extended
  # wane_imm_fract_weekly_3_logit = par[4]
  # Imm_3_partial
  df_all  = df_Rt_value_pop
  df_all = df_all[1:days_to_week$n_full_weeks,]
  # weeks and days
  nweeks = days_to_week$nweeks
  weekly_daygroups_t = days_to_week$weekly_daygroups_t
  weekly_daygroups_tlast = days_to_week$weekly_daygroups_tlast
  
  ## transformations
  Imm_2_partial = inv_logit(Imm_2_partial_logit)
  wane_imm_fract_weekly_1 = inv_logit(wane_imm_fract_weekly_1_logit) # scale daily/weekly
  wane_imm_fract_weekly_2 = inv_logit(wane_imm_fract_weekly_2_logit) # scale daily/weekly
  # wane_imm_fract_weekly_3 = exp(wane_imm_fract_weekly_3_log)
  
  ## country loop
  # for each location, simulate immunity and update target based on conditions
  location_v = unique(df_all$location_name)
  df_save = list()
  target = 0 # initiate target for optimisation
  min_target = 0 # initiate minimum target for optimisation
  for (i in location_v) { # i = location_v[1]
    df_i = df_all %>% filter(location_name==i) 
    
    #### value: infection intensity
    if (!simulate) {
      # normalise infection intensity (so that infection attack rate is 1% each year)
      laste_year_ii = mlast(df_i$value,n=365)/mlast(df_i$population,n=365)
      df_i$value = df_i$value * ( 0.01/sum(  laste_year_ii  )  )
    }
    if (simulate) {
      df_i$value = 0; df_i$value[ weekly_daygroups_tlast[[1]][1] ] = sim_o$inf_ini
    }
    
    #### Imm_X: the different immunity compartments
    df_i$Imm_1 = NA; 
    df_i$Imm_2 = NA; 
    if (!simulate) {
      # initial Rt and immunity
      Rt_ini = mean(df_i$Rt[1:p$Rt_for_ini_est]) # avoiding impact of noise in Rt 
      Imm_ini = (p$R0 - Rt_ini)/p$R0 # back-calculating immunity from Rt and R0
      # assuming that all immunity is recent (thus all into first immunity compartment)
      df_i$Imm_1[weekly_daygroups_tlast[[1]]] = Imm_ini # full immunity
      df_i$Imm_2[weekly_daygroups_tlast[[1]]] = 0 # partial immunity
    }
    if (simulate) {
      # initial immunity
      df_i$Imm_1[weekly_daygroups_tlast[[1]]] = sim_o$Imm_1_ini
      df_i$Imm_2[weekly_daygroups_tlast[[1]]] = sim_o$Imm_2_ini
    }
    # build immunity from the immunity compartments
    df_i$Imm = NA; 
    df_i$Imm[weekly_daygroups_tlast[[1]]] = df_i$Imm_1[weekly_daygroups_tlast[[1]]] + df_i$Imm_2[weekly_daygroups_tlast[[1]]]*Imm_2_partial # immunity combined
    
    #### waning_X: waning from the different immunity compartments
    # initiate waning
    df_i$waning_1 = NA; df_i$waning_1[weekly_daygroups_tlast[[1]]] = 0
    df_i$waning_2 = NA; df_i$waning_2[weekly_daygroups_tlast[[1]]] = 0
    df_i$new_Imm = NA; df_i$new_Imm[weekly_daygroups_tlast[[1]]] = 0
    
    #### Rt_without_immunity and Rt with immmunity
    if (!simulate){
      # compute Rt without immunity
      df_i$Rt_without_immunity = NA; 
      df_i$Rt_without_immunity[weekly_daygroups_tlast[[1]]] = df_i$Rt[weekly_daygroups_tlast[[1]]] * 1/(1-Imm_ini)
    }
    if (simulate) {
      # Rt without immunity exist, compute Rt effective
      df_i$Rt_eff = NA
      df_i$Rt_without_immunity = df_i$Rt_seasonal 
      df_i$Rt_eff[weekly_daygroups_tlast[[1]]] = 
        df_i$Rt_without_immunity[weekly_daygroups_tlast[[1]]] * (1-df_i$Imm[weekly_daygroups_tlast[[1]]])
    }
    
    #### sX: different scenarios
    ## counter-factuals/scenarios/helpers
    df_i$s1 = NA; df_i$s1[weekly_daygroups_tlast[[1]]] = 1
    df_i$s1_1 = NA; df_i$s1_1[weekly_daygroups_tlast[[1]]] = 1
    df_i$s1_2 = NA; df_i$s1_2[weekly_daygroups_tlast[[1]]] = 0
    # 
    df_i$s2 = NA; df_i$s2[weekly_daygroups_tlast[[1]]] = 1
    #
    df_i$s3 = NA; df_i$s3[weekly_daygroups_tlast[[1]]] = df_i$s1[weekly_daygroups_tlast[[1]]]*df_i$Imm[weekly_daygroups_tlast[[1]]]
    
    # time loop (through chunks of days of a given location)
    for (t in 1:nweeks ) { # t = 1
      # deal with weekly to daily
      t_days  = weekly_daygroups_t[[t]]
      tlast_days  = weekly_daygroups_tlast[[t]]
      # simlate infections (value) daily stemming from last week 
      if (simulate) {
        # handle daily growth simulation 1 day-chunk at a time
        for (chunk_day_i in 1:length(tlast_days) ){ # chunk_day_i = 1
          whichday = tlast_days[chunk_day_i]
          current_infection = df_i$value[whichday]
          current_Rt = df_i$Rt_eff[whichday]
          current_offspring = current_infection * current_Rt
          
          si_v_along = c(1:length(sim_o$si_distribution))
          
          ## add offspring incident infections back into value
          distributed_offspring = current_offspring * sim_o$si_distribution
          # truncate if offspring goes beyond 
          n_too_long = ( whichday+mlast(si_v_along) ) - days_to_week$n_full_weeks
          trunk_v = c( 1:( length(whichday+si_v_along)-max(n_too_long,0) ) ) 
          # add the offspring
          df_i$value[(whichday+si_v_along)[trunk_v]] = df_i$value[(whichday+si_v_along)[trunk_v]] + 
            distributed_offspring[trunk_v]
        } # df_i$value[1: (whichday+mlast(si_v_along)) ] %>% plot()
      }
      
      # immunity waning (allowing different immunity compartments)
      # Imm_1 is the full immunity compartment
      # Imm_2 3, ... are partial immunity compartments
      waning_1 = df_i$Imm_1[tlast_days[1]]*wane_imm_fract_weekly_1 
      waning_2 = df_i$Imm_2[tlast_days[1]]*wane_imm_fract_weekly_2 
      df_i$waning_1[t_days] = waning_1/7 # distribute over days
      df_i$waning_2[t_days] = waning_2/7 # distribute over days
      # accumulation of new immunity (proportional to: value = "infection intensity",dark_factor,NE)
      new_infections = sum(df_i$value[tlast_days]/df_i$population[tlast_days])
      new_Imm = new_infections * p$ar_and_re
      df_i$new_Imm[t_days] = new_Imm/7 # distribute over days
      # new infections also come from partly susceptible Imm compartments
      full_susc = 1 - ( df_i$Imm_1[tlast_days[1]] + df_i$Imm_2[tlast_days[1]] )
      if (full_susc<0) print("warning0")
      new_Imm_from_Imm_2 = df_i$Imm_2[tlast_days[1]]*Imm_2_partial / 
        ( df_i$Imm_2[tlast_days[1]]*Imm_2_partial + full_susc*1 )
      # apply changes to Immunity compartments
      df_i$Imm_1[t_days] = df_i$Imm_1[tlast_days[1]] - waning_1 + new_Imm
      df_i$Imm_2[t_days] = df_i$Imm_2[tlast_days[1]] - waning_2 + waning_1 - new_Imm_from_Imm_2*new_Imm
      if (any(df_i$Imm_1[t_days]<0)) print("warning1")
      if (any(df_i$Imm_2[t_days]<0)) {print("warning2");}
      # recompute combined "effective" immunity
      df_i$Imm[t_days] = df_i$Imm_1[t_days] + df_i$Imm_2[t_days]*Imm_2_partial
      # compute new Rt without immunity (Rt AS IF immunity would have vanished)
      if (!simulate) df_i$Rt_without_immunity[t_days] = df_i$Rt[t_days] * 1/(1-df_i$Imm[t_days])
      if ( simulate) df_i$Rt_eff[t_days] = df_i$Rt_without_immunity[t_days] * (1-df_i$Imm[t_days])
      
      ## scenarios 
      # 1: just waning shape
      waning_s1_1 = df_i$s1_1[tlast_days[1]]*wane_imm_fract_weekly_1 
      waning_s1_2 = df_i$s1_2[tlast_days[1]]*wane_imm_fract_weekly_2
      df_i$s1_1[t_days] = df_i$s1_1[tlast_days] - waning_s1_1 
      df_i$s1_2[t_days] = df_i$s1_2[tlast_days] - waning_s1_2 + waning_s1_1
      df_i$s1[t_days] = df_i$s1_1[t_days] + df_i$s1_2[t_days]*Imm_2_partial
      # 2: just accounting for new immunity
      new_week_eff_immunity = sum(df_i$value[tlast_days]/df_i$population[tlast_days]) * p$ar_and_re
      new_week_Rt_modifier = 1 - new_week_eff_immunity
      new_week_Rt_modifier_inv = 1 / new_week_Rt_modifier
      df_i$s2[t_days] = df_i$s2[tlast_days] * new_week_Rt_modifier_inv
      # 3: just accounting for immunity from the start
      df_i$s3[t_days] = df_i$s1[t_days]*df_i$Imm[1]
    }
    
    if (!simulate){
      ## "target" allows multiple constraints
      # regularize: immunity last year ~ immunity 1 year later
      Imm_diff = df_i$Imm[df_i$Rt_at_one==1] - df_i$Imm[df_i$Rt_at_one==2]
      target_Imm_diff = -dnorm(x=Imm_diff,mean = 0,sd=0.05, log = T)
      target_Imm_diff_min = -dnorm(x=0,mean = 0,sd=0.05, log = T)
      # regularize: R0 based on Rt and R0
      # note: Rt_without immunity can exceed R0, since Rt and Immunity beyong the initial condition don't know about each other 
      max_Rt_without_immunity = max(df_i$Rt_without_immunity,na.rm=T)
      target_Rt_without_immunity = -dnorm(x=max_Rt_without_immunity,mean = p$R0,sd=1, log = T)
      target_Rt_without_immunity_min = -dnorm(x=p$R0,mean = p$R0,sd=1, log = T)
      # regularize; effective waning 
      
      # could also regularize; max immunity in the population
      
      target = target + target_Imm_diff + target_Rt_without_immunity
      min_target = min_target = target_Imm_diff_min + target_Rt_without_immunity_min # -1.157855
    }
    
    # print(paste("min target:",min_target %>% round(2)))
    # print(paste("Max new_Imm",max(df_i$new_Imm) %>% round(3),i))
    # print(paste("Average waning",mean(df_i$waning) %>% round(3),i))
    df_save[[i]] = df_i
  }
  # pack it all up
  df_return = bind_rows(df_save) %>% 
    mutate(target=target, 
           min_target=min_target)
  
  # return depending on where the function is used
  if (for_optim==T) return(target)
  if (for_optim==F) return(df_return)
}
```

### *"Infection intensity" data*

```{r infection_intensity, paged.print=TRUE}
# date, value, location
path_data="../../../Big input data/2024-01-17_dk_wastewater_data.csv"
x=read_csv(path_data)
x %>% filter(date>"2022-10-01",date<"2024-02-03") %>% ggplot(aes(date,rna_mean_faeces)) + geom_line() # 
# ready for main
df_ii_weekly = x %>% select(date=date,
                            value=rna_mean_faeces) %>% 
  mutate(location="DK")
no_duplication_no_gaps_right_order(df_ii_weekly)
df_ii = weekly_make_daily(df_ii_weekly)
df_ii= df_ii %>% fill_log(value, .direction="up")
no_duplication_no_gaps_right_order(df_ii)
range_df_ii = range(df_ii$date)
```

### *Obtaining Rt*

```{r transmission_intensity}
library(EpiEstim) # EpiEstim as simple choice to get Rt
n_location = n_distinct(df_ii$location)
df_country = list()
for (i in 1:n_location){
  country_i = unique(df_ii$location)[i]
  case_incidenc_location = df_ii %>% filter(location ==country_i) 
  ## estimate slope
  case_incidenc_location$value_slope = NA
  for (i in 2:nrow(case_incidenc_location) ) {
    last_x = case_incidenc_location$value[i-1]
    curr_x = case_incidenc_location$value[i]
    case_incidenc_location$value_slope[i] = ( curr_x/last_x )
  }
  ## estimate Rt
  rt <- estimate_R(case_incidenc_location$value*1000, 
                   method="parametric_si",
                   config = make_config(list(
                     mean_si = p$si_mean, 
                     std_si = p$si_sd) ) )
  # extract 
  filtered_Rt = rt$R$`Mean(R)`
  filtered_Rt[filtered_Rt>5] = NA # replacing extreme Rt values
  filtered_Rt[filtered_Rt<0.2] = NA # replacing extreme Rt values
  Rt_df = tibble( id=rt$R$t_start,Rt=filtered_Rt )
  
  df_country[[i]] = case_incidenc_location %>% 
    mutate(id=c(1:n())) %>% left_join(Rt_df,by="id") %>% select(-id)
}
df_transmission = bind_rows(df_country)
check_df_transmission(df_transmission) 
```

```{r}
Rt_one_1 = ymd("2022-12-22")
Rt_one_2 = ymd("2023-12-15")

df_transmission %>% filter(date>"2022-06-01",date<"2024-02-03") %>%  ggplot(aes(x=date,y=value))+ 
  geom_line() +
  geom_vline(xintercept = ymd(c("2022-12-08","2023-11-23")),linetype="dashed")
df_transmission %>% filter(date>"2022-10-01",date<"2024-02-03") %>% ggplot(aes(x=date,y=Rt))+ 
  geom_line() + 
  geom_hline(yintercept = 1,linetype="dashed")+
  geom_vline(xintercept = c(Rt_one_1,Rt_one_2),linetype="dashed")



df_transmission %>% filter(date>"2022-10-01",date<"2024-02-03") %>% ggplot()+ 
  geom_line(aes(x=date,
                y=value_slope ) ) + 
  geom_line(aes(x=date,
                y=Rt ),col="red" ) + 
  geom_hline(yintercept = 1,linetype="dashed")+
  geom_vline(xintercept = c(Rt_one_1,Rt_one_2),linetype="dashed")
```

### *Modelling immunity against transmission and obtain augmented Rt*

```{r echo=TRUE}
# add poplation sizes
df_all = df_transmission %>% left_join(aux$pops %>% select(location,location_name,population),by = join_by(location)) 
transmission_focus_dates = (df_all$date) %>% range()

df_all$Rt_at_one = 0
df_all$Rt_at_one[df_all$date == Rt_one_1] = 1
df_all$Rt_at_one[df_all$date == Rt_one_2] = 2

df_all$date[df_all$Rt_at_one!=0]
# df should start when Rt ~ 1
df_Rt_value_pop = df_all %>% filter_log(date>=Rt_one_1)
not_too_short_or_long(df_Rt_value_pop)
```

```{r optimising_boe_parameters}
# optimising immunity_boe function
days_to_week=get_weekly_daygroups(df_Rt_value_pop)

f = function(...) compute_Rt_without_immunity(...,
                                              df_Rt_value_pop=df_Rt_value_pop,
                                              for_optim = T,
                                              p=p,days_to_week=days_to_week)
grid_i = 1
df_grid = list()
n_grid = 20
n_grid^4
###############
# the following rule does not really work for me, even low X% leas to faster halving
# rule of thumb: something that increases with X% each week, takes 72/X weeks to double

# 152.27   0.05   0.17   0.05
if (T) { # 4 mins
  tic()
  for (NE_and_dark_factor_i in seq(from=log(80),to=log(140),length.out=15) ) {
    for ( wane_imm_fract_weekly_1_log_i in seq(from=logit(0.05),to=logit(0.05),length.out=1) ) {
      for ( Imm_2_partial_i in seq(from=logit(0.1),to=logit(0.9),length.out=20) ) {
        for ( wane_imm_fract_weekly_2_log_i in seq(from=logit(0.06),to=logit(0.06),length.out=1) ) {
          
          #######################################
          target_i = f( par=c(NE_and_dark_factor_i,
                              wane_imm_fract_weekly_1_log_i,
                              Imm_2_partial_i,
                              wane_imm_fract_weekly_2_log_i) 
          )
          df_grid[[grid_i]] = tibble(NE_and_dark_factor=NE_and_dark_factor_i,
                                     wane_imm_fract_weekly_1_log=wane_imm_fract_weekly_1_log_i,
                                     Imm_2_partial = Imm_2_partial_i,
                                     wane_imm_fract_weekly_2_log=wane_imm_fract_weekly_2_log_i,
                                     target=target_i)
          grid_i = grid_i + 1
          #########################################
        }
        
      }
    }
  }
  toc()
}

#################
df_grid_all = bind_rows(df_grid)
df_grid_all %>% ggplot(aes(x=NE_and_dark_factor,y=target)) + geom_point() + scale_y_log10()
df_grid_all %>% ggplot(aes(x=wane_imm_fract_weekly_1_log,y=target)) + geom_point()+ scale_y_log10()
df_grid_all %>% ggplot(aes(x=Imm_2_partial,y=target)) + geom_point()+ scale_y_log10()
df_grid_all %>% ggplot(aes(x=wane_imm_fract_weekly_2_log,y=target)) + geom_point()+ scale_y_log10()

df_min = df_grid_all %>% filter(target==min(target)) 
df_Rt_without_imm = compute_Rt_without_immunity(par=c(df_min$NE_and_dark_factor,
                                                      df_min$wane_imm_fract_weekly_1_log,
                                                      df_min$Imm_2_partial,
                                                      df_min$wane_imm_fract_weekly_2_log) , 
                                                df_Rt_value_pop=df_Rt_value_pop , 
                                                for_optim=F,
                                                p=p,days_to_week=days_to_week)

par_min = c(df_min$NE_and_dark_factor,
            df_min$wane_imm_fract_weekly_1_log,
            df_min$Imm_2_partial,
            df_min$wane_imm_fract_weekly_2_log) 

par_min %>% unscale_pars() %>% round(2)

df_Rt_without_imm = compute_Rt_without_immunity(par_min,
                                                df_Rt_value_pop=df_Rt_value_pop,
                                                for_optim = F,
                                                p=p,days_to_week=days_to_week)

# 120.30   0.15   0.40   0.00


```

```{r}
par_optim = par_min
if (F){
  f = function(...) compute_Rt_without_immunity(...,
                                                wane_imm_fract_weekly_1_log = par_min[2], # fix
                                                Imm_2_partial=par_min[3], # fix
                                                wane_imm_fract_weekly_2_log = par_min[4], # fix
                                                df_Rt_value_pop=df_Rt_value_pop,
                                                for_optim = T,
                                                p=p,days_to_week=days_to_week)
  res_optim = optim(par=par_min[1],fn=f)
  
  # previously: 
  par_min %>% unscale_pars() %>% round(2)
  res_optim$par %>% unscale_pars() %>% round(2)
  
  df_Rt_without_imm = compute_Rt_without_immunity(par=res_optim$par,
                                                  Imm_2_partial=par_min[3], # fix
                                                  wane_imm_fract_weekly_2_log = par_min[4], # fix
                                                  df_Rt_value_pop,
                                                  for_optim = F,
                                                  p=p,days_to_week=days_to_week)
}

```

```{r}
# look
date_start = "2022-11-12"

tibble(
  date_range=c("ii_range","transmission_focus","Rt_one"),
  date_lower=c(range_df_ii[1],transmission_focus_dates[1],Rt_one_1),
  date_upper=c(range_df_ii[2],transmission_focus_dates[2],Rt_one_2)) %>% 
  ggplot(aes(y=date_range,yend=date_range)) +
  geom_segment(aes(x=date_lower,xend=date_upper))

df_Rt_without_imm %>% 
  filter(date>date_start) %>% 
  ggplot(aes(x=date)) + 
  geom_line(aes(y=s1),col="orange") + 
  facet_wrap(~location,scales="free_y") + coord_cartesian(ylim=c(0,1)) + 
  theme(axis.ticks.y=element_blank()) + labs(subtitle = "Just the waning function")

df_Rt_without_imm %>% 
  filter(date>date_start) %>% 
  ggplot(aes(x=date)) + 
  geom_line(aes(y=s3),col="orange") + 
  geom_line(aes(y=1-1/s2),col="blue") + 
  geom_line(aes(y=Imm),col="black",linetype="dashed") +
  geom_vline(xintercept = c(Rt_one_1,Rt_one_2),linetype="dashed") +
  facet_wrap(~location,scales="free_y") + 
  theme(axis.ticks.y=element_blank()) + 
  labs(y="",x="",subtitle = "Immunity from last year's bulk burden (orange) + new immunity (blue)")

df_Rt_without_imm %>% 
  filter(date>date_start) %>% 
  ggplot(aes(x=date)) + 
  geom_line(aes(y=Rt),col="black") + 
  geom_hline(yintercept = 1,linetype="dashed") + facet_wrap(~location,scales="free_y") + 
  theme(axis.ticks.y=element_blank()) + labs(subtitle = "Rt empirical, estimated from infection intensity indicator")
df_Rt_without_imm %>% 
  filter(date>date_start) %>% 
  ggplot(aes(x=date)) + 
  geom_line(aes(y=Rt),col="black") + 
  geom_line(aes(y=s2*Rt),col="blue") + 
  geom_vline(xintercept = c(Rt_one_1,Rt_one_2),linetype="dashed")+
  geom_hline(yintercept = 1,linetype="dashed") + facet_wrap(~location,scales="free_y") + 
  theme(axis.ticks.y=element_blank()) + labs(subtitle = "Rt empirical (black) and accounting for new immunity only (blue)")

df_Rt_without_imm %>% 
  filter(date>date_start) %>% 
  ggplot(aes(x=date)) + 
  geom_line(aes(y=Rt_without_immunity),col="purple") + 
  geom_vline(xintercept = c(Rt_one_1,Rt_one_2),linetype="dashed")+
  geom_hline(yintercept = 1,linetype="dashed") + facet_wrap(~location,scales="free_y") + 
  theme(axis.ticks.y=element_blank()) + labs(subtitle = "Rt as if no past or new immunity")


```

### *Fit augmented Rt*

```{r fig.height=10, fig.width=10, message=TRUE}

# ---- |-Prepare df for stan ----
df_stan = df_Rt_without_imm %>% ungroup() %>% 
  filter(!is.na(Rt)) %>% 
  mutate(season_t=as.numeric(date-ymd("2022-12-31")) ) %>% 
  mutate(id=c(1:n()))

# ---- |-Create stan list ----
stan_list = list(
  # model input related to data-points, country structure, etc
  n = nrow(df_stan),
  n_location = n_distinct(df_stan$location),
  location_id = fct_inorder(df_stan$location) %>% as.numeric(),
  location_id_raw = fct_inorder(df_stan$location) %>% levels() %>% enframe,
  # model input related to seasonality and outcome
  season_t = df_stan$season_t,
  y = df_stan$Rt_without_immunity,
  # model input related to prior information
  prior = 2
)

checking_stan_list(stan_list)

# ---- |-Create stan list ----
stan_refit = F
if (stan_refit) {
  fit_season=rstan::stan(
  file="stan/season00_start.stan",
  chains=8 ,thin=8,iter=800,# 1200,
  seed=12, cores = getOption("mc.cores", 1L),
  control=list(
    #adapt_delta=0.9,
    # max_treedepth=14
  ),
  data=stan_list
) # 
save(fit_season,stan_list,file="../Big data/fit_season.Rdata")
}

load(file="../Big data/fit_season.Rdata")
fit = fit_season

library(rethinking)
precis(fit,pars=c("season_fold","season_amp"),depth=2) # 40-50% more beta in winter compared to summer
precis(fit,pars=c("season_fold_log_mu","season_fold_log_sd","y_base","sigma"),depth=2)

mcmc_areas(fit,pars="season_fold[1]") #+ coord_cartesian(xlim=c(1,1.75))

# ---- |-Predictive checks ----
df_pp = fit %>% gather_draws(mu[id]) %>% filter(.draw%in%c(1:50)) %>% 
  left_join(df_stan , by="id")

library(ggtext)
st = "Transmission in the absence of immunity (black) after removing the effect of immunity, as well as fitted seasonality (blue)"
df_stan %>% 
  ggplot(aes(x=date,y=Rt_without_immunity)) + 
  geom_line() + geom_hline(yintercept = 1) + 
  geom_line(data=df_pp,aes(x=date,y=.value,group=.draw),alpha=0.2,col="blue" ) +
  facet_wrap(~location,scales="free_y") + 
  labs(subtitle=st)+
  theme(axis.ticks.y=element_blank(),
        plot.subtitle = element_textbox_simple(valign = 1))
```

### *Simulate forward*

We finally simulate 1 year ahead. The season Rt looks reasonable. There is an issue with too much immunity (warning0) and there is an issue with infections disappearing to zero before they can be picked up again by increasing Rt.

```{r}
if (F){
  # simulate forward
  
  # take 1 seasonal Rt curve
  Rt_seasonal = fit %>% gather_draws(mu[id]) %>% ungroup() %>% 
    filter(.draw%in%c(1)) %>% 
    left_join(df_stan , by="id") %>% 
    select(date,Rt_seasonal=.value,population,location_name) %>% 
    # to simulate, shift the dates by 1 year
    mutate( date=date+365 ) %>% 
    mutate( Rt_at_one=0 ) 
  Rt_seasonal$Rt_at_one[1] = 1
  Rt_seasonal$Rt_at_one[365] = 2
  
  
  # get the week-day details
  days_to_week_sim = get_weekly_daygroups(Rt_seasonal)
  
  # run compute_Rt_without_immunity with diff input
  # give counterfactual Rt
  # give intial immunity
  Imm_1_ini = df_Rt_without_imm$Imm_1[df_Rt_without_imm$Rt_at_one==2]
  Imm_2_ini = df_Rt_without_imm$Imm_2[df_Rt_without_imm$Rt_at_one==2]
  
  df_Rt_without_imm = compute_Rt_without_immunity(par=par_optim,
                                                  df_Rt_value_pop=Rt_seasonal,
                                                  for_optim = F,
                                                  p=p,
                                                  days_to_week=days_to_week_sim,
                                                  sim_o = list(
                                                    inf_ini = 1000*4,
                                                    Imm_1_ini = Imm_1_ini,
                                                    Imm_2_ini = Imm_2_ini,
                                                    si_distribution = p$si_discr_dist
                                                  ) )
  df_Rt_without_imm %>% g()
  df_Rt_without_imm$Rt_seasonal %>% plot()
  df_Rt_without_imm$Rt_eff %>% plot()
  df_Rt_without_imm$Imm %>% plot()
  
  df_Rt_without_imm %>% ggplot(aes(x=date)) +
    geom_line(aes(y=value))
}

```
