---
output: html_document
---

# Season BOE: A Simple COVID-19 Back-of-the-Envelope Model

## Introduction

-   Given that SARS-CoV2 immunity as reached 'equilibrium', can we model viral spread in a very simplified way?

-   high viral attack rate (likely above 50% per year) relative to typical speed of waning (substantial waning within 1/2 year)

    -   immunity has reached an equilibrium: majority of individuals in EU/EEA have had multiple exposures and vaccines, and for impact of infection on transmission the immunity history does not matter

        -   Thus, infection impact on population immunity can be modelled similarity for 22/23 as for 23/24

\*absence of NPIs, and less prevalence-elastic behaviour -\> Rt mostly impacted by immunity, seasonality, variants

\*by accounting for immunity (and variants) can we estimate seasonality?

## Method assumptions in very brief

-   Infection-intensity indicator is *proportional* to the true infection incidence (clearly NOT true with reported cases)

-   Viral transmission is *only* driven through seasonality, immunity accumulation and waning (and new variants)

    -   Thus, at a similar time of the year, a similar Rt value implies a similar level of prevailing population-immunity (assuming no impact of variants)

    -   For example we could say: in Nov 2022 and in Nov 2023 when Rt was close to 1, population-immunity was nearly the same

-   Immunity waning occurs in a relatively simple way ( 100%-immunity-compartment -\> X%-immunity-compartment -\> susceptible with exponential flow )

-   Parametric assumptions

    -   Serial interval (si)=7.7 days

    -   R0 (r0) = 6

    -   Yearly infection attack-rate (ar): 60%

    -   Reduction in susceptibility through infection (re): 90%

## Conceptual model walk through

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	include = FALSE
)
# set R up 
source("../setup.R")
source("settings.R")
source("checks.R")
```

```{r}
compute_Rt_without_immunity = function( par, 
                                        df_Rt_value_pop,
                                        for_optim=F,
                                        p) {
  ## recasting parameters (needed due to optim use)
  p$ar_and_re = par[1]
  wane_imm_fract_weekly_1_log = par[2]
  Imm_2_partial = par[3]
  wane_imm_fract_weekly_2_log = par[4]
  # waning compartments can be extended
  # wane_imm_fract_weekly_3_log = par[4]
  # Imm_3_partial
  df_all  = df_Rt_value_pop
  
  ## transformations
  wane_imm_fract_weekly_1 = exp(wane_imm_fract_weekly_1_log)
  wane_imm_fract_weekly_2 = exp(wane_imm_fract_weekly_2_log)
  # wane_imm_fract_weekly_3 = exp(wane_imm_fract_weekly_3_log)
  
  ## country loop
  # for each location, simulate immunity and update target based on conditions
  location_v = unique(df_all$location_name)
  df_save = list()
  target = 0 # initiate target for optimisation
  min_target = 0 # initiate minimum target for optimisation
  for (i in location_v) { # i = location_v[1]
    df_i = df_all %>% filter(location_name==i) 
    # normalise infection intensity
    df_i$value = df_i$value * (0.01/sum( (df_i$value/df_i$population) , na.rm=T  )  )
    # if(sum(df_i$value/df_i$population) != 0.01) print(paste("Check normalising inc for",i))
    
    # initial state
    Rt_ini = mean(df_i$Rt[1:p$Rt_for_ini_est]) # avoiding impact of noise in Rt 
    Imm_ini = (p$R0 - Rt_ini)/p$R0 # back-calculating immunity from Rt and R0
    # assuming that all immunity is recent (thus all into first immunity compartment)
    df_i$Imm_1 = NA; df_i$Imm_1[1] = Imm_ini # full immunity
    df_i$Imm_2 = NA; df_i$Imm_2[1] = 0 # partial immunity
    df_i$Imm = NA; df_i$Imm[1] = df_i$Imm_1[1] + df_i$Imm_2[1]*Imm_2_partial # immunity combined
    #
    df_i$waning_1 = NA; df_i$waning_1[1] = 0
    df_i$waning_2 = NA; df_i$waning_2[1] = 0
    df_i$new_Imm = NA; df_i$new_Imm[1] = 0
    df_i$Rt_without_immunity = NA; 
    df_i$Rt_without_immunity[1] = df_i$Rt[1] * 1/(1-Imm_ini)
    
    ## counter-factuals/scenarios/helpers
    df_i$s1 = NA; df_i$s1[1] = 1
    df_i$s1_1 = NA; df_i$s1_1[1] = 1
    df_i$s1_2 = NA; df_i$s1_2[1] = 0
    # 
    df_i$s2 = NA; df_i$s2[1] = 1
    #
    df_i$s3 = NA; df_i$s3[1] = df_i$s1[1]*df_i$Imm[1]
    
    # time loop (through the weeks of a given location)
    for (t in 2:nrow(df_i) ) { # t = 2
      # immunity waning (allowing different immunity compartments)
      # Imm_1 is the full immunity compartment
      # Imm_2 3, ... are partial immunity compartments
      waning_1 = df_i$Imm_1[t-1]*wane_imm_fract_weekly_1 
      waning_2 = df_i$Imm_2[t-1]*wane_imm_fract_weekly_2 
      df_i$waning_1[t] = waning_1
      df_i$waning_2[t] = waning_2
      # accumulation of new immunity (proportional to: value = "infection intensity",dark_factor,NE)
      new_Imm = df_i$value[t-1]/df_i$population[t-1] * p$ar_and_re
      df_i$new_Imm[t] = new_Imm
      # apply changes to Immunity compartments
      df_i$Imm_1[t] = df_i$Imm_1[t-1] - waning_1 + new_Imm
      df_i$Imm_2[t] = df_i$Imm_2[t-1] - waning_2 + waning_1
      if (df_i$Imm_1[t]<0) print("warning1")
      if (df_i$Imm_2[t]<0) print("warning2")
      # recompute combined "effective" immunity
      df_i$Imm[t] = df_i$Imm_1[t] + df_i$Imm_2[t]*Imm_2_partial
      # compute new Rt without immunity (Rt AS IF immunity would have vanished)
      df_i$Rt_without_immunity[t] = df_i$Rt[t] * 1/(1-df_i$Imm[t])
      
      ## scenarios 
      # 1: just waning shape
      waning_s1_1 = df_i$s1_1[t-1]*wane_imm_fract_weekly_1 
      waning_s1_2 = df_i$s1_2[t-1]*wane_imm_fract_weekly_2
      df_i$s1_1[t] = df_i$s1_1[t-1] - waning_s1_1 
      df_i$s1_2[t] = df_i$s1_2[t-1] - waning_s1_2 + waning_s1_1
      df_i$s1[t] = df_i$s1_1[t] + df_i$s1_2[t]*Imm_2_partial
      # 2: just accounting for new immunity
      new_week_eff_immunity = df_i$value[t-1]/df_i$population[t-1] * p$ar_and_re
      new_week_Rt_modifier = 1 - new_week_eff_immunity
      new_week_Rt_modifier_inv = 1 / new_week_Rt_modifier
      df_i$s2[t] = df_i$s2[t-1] * new_week_Rt_modifier_inv
      # 3: just accounting for immunity from the start
      df_i$s3[t] = df_i$s1[t]*df_i$Imm[1]
    }
    
    ## "target" allows multiple constraints
    # regularize: immunity last year ~ immunity 1 year later
    Imm_diff = df_i$Imm[df_i$Rt_at_zero==1] - df_i$Imm[df_i$Rt_at_zero==2]
    target_Imm_diff = -dnorm(x=Imm_diff,mean = 0,sd=0.05, log = T)
    target_Imm_diff_min = -dnorm(x=0,mean = 0,sd=0.05, log = T)
    # regularize: R0 based on Rt and R0
    # note: Rt_without immunity can exceed R0, since Rt and Immunity beyong the initial condition don't know about each other 
    max_Rt_without_immunity = max(df_i$Rt_without_immunity,na.rm=T)
    target_Rt_without_immunity = -dnorm(x=max_Rt_without_immunity,mean = p$R0,sd=1, log = T)
    target_Rt_without_immunity_min = -dnorm(x=p$R0,mean = p$R0,sd=1, log = T)
    # regularize; max immunity in the population
    
    target = target + target_Imm_diff + target_Rt_without_immunity
    min_target = min_target = target_Imm_diff_min + target_Rt_without_immunity_min # -1.157855
    # print(paste("min target:",min_target %>% round(2)))
    # print(paste("Max new_Imm",max(df_i$new_Imm) %>% round(3),i))
    # print(paste("Average waning",mean(df_i$waning) %>% round(3),i))
    df_save[[i]] = df_i
  }
  df_return = bind_rows(df_save) %>% 
    mutate(target=target, 
           min_target=min_target)
  
  if (for_optim==T) return(target)
  if (for_optim==F) return(df_return)
}
```

### "Infection intensity" data

```{r infection_intensity, paged.print=TRUE}
# date, value, location
path_data="../../../Big input data/data_positive_tests.csv"
x=read.csv(path_data,sep = ";",dec = ",") %>% mutate(WEEK=ymd(WEEK))
x = x %>% select(date=WEEK,
             value=POS_TEST_PERC) %>% 
  mutate(location="NL")
df_ii = x
no_duplication_no_gaps_right_order(df_ii)
df_ii = weekly_make_daily(df_ii)
df_ii= df_ii %>% fill_log(value, .direction="up")
no_duplication_no_gaps_right_order(df_ii)
range_df_ii = range(df_ii$date)
```

### Obtaining Rt

```{r transmission_intensity}
library(EpiEstim) # EpiEstim as simple choice to get Rt
n_location = n_distinct(df_ii$location)
df_country = list()
for (i in 1:n_location){
  country_i = unique(df_ii$location)[i]
  case_incidenc_location = df_ii %>% filter(location ==country_i) 
  ## estimate slope
  case_incidenc_location$value_slope = NA
  for (i in 2:nrow(case_incidenc_location) ) {
    last_x = case_incidenc_location$value[i-1]
    curr_x = case_incidenc_location$value[i]
    case_incidenc_location$value_slope[i] = ( curr_x/last_x )
  }
  ## estimate Rt
  rt <- estimate_R(case_incidenc_location$value*1000, 
                   method="parametric_si",
                   config = make_config(list(
                     mean_si = p$si_mean, 
                     std_si = p$si_sd) ) )
  # extract 
  filtered_Rt = rt$R$`Mean(R)`
  filtered_Rt[filtered_Rt>5] = NA # replacing extreme Rt values
  filtered_Rt[filtered_Rt<0.2] = NA # replacing extreme Rt values
  Rt_df = tibble( id=rt$R$t_start,Rt=filtered_Rt )
  
  df_country[[i]] = case_incidenc_location %>% 
    mutate(id=c(1:n())) %>% left_join(Rt_df,by="id") %>% select(-id)
}
df_transmission = bind_rows(df_country)
check_df_transmission(df_transmission) 
```

```{r}
df_transmission %>% filter(date>"2022-06-01",date<"2024-02-03") %>%  ggplot(aes(x=date,y=value))+ 
  geom_line() +
  geom_vline(xintercept = ymd(c("2022-12-08","2023-11-23")),linetype="dashed")
df_transmission %>% filter(date>"2022-10-01",date<"2024-02-03") %>% ggplot(aes(x=date,y=Rt))+ 
  geom_line() + 
  geom_hline(yintercept = 1,linetype="dashed")+
  geom_vline(xintercept = ymd(c("2022-12-08","2023-11-23")),linetype="dashed")

df_transmission %>% filter(date>"2022-10-01",date<"2024-02-03") %>% ggplot()+ 
  geom_line(aes(x=date,
                y=(value_slope-min(value_slope))/(max(value_slope)-min(value_slope)) ) ) + 
  geom_line(aes(x=date,y= (Rt-min(Rt)) / (max(Rt)-min(Rt))  ),col="orange" ) + 
  geom_hline(yintercept = 1,linetype="dashed")+
  geom_vline(xintercept = ymd(c("2022-12-08","2023-11-23")),linetype="dashed")
```

### Modelling immunity against transmission and obtain augmented Rt

```{r echo=TRUE}
# add poplation sizes
df_all = df_transmission %>% left_join(aux$pops,by = join_by(location)) %>% filter(date>"2022-10-01",date<"2024-02-03")
transmission_focus_dates = (df_all$date) %>% range()

df_all$population =  pop_myexample
df_all$location_name = "EU"
df_all$Rt_at_zero = 0
df_all$Rt_at_zero[10] = 1 # TODO remove hard-coding
df_all$Rt_at_zero[60] = 2 # TODO remove hard-coding
(Rt_zero_1 = df_all$date[10]) # "2022-11-12"
(Rt_zero_2 = df_all$date[60]) # "2023-11-23"

# df should start when Rt ~ 1
df_Rt_value_pop = df_all %>% filter_log(date>=Rt_zero_1)
not_too_short_or_long(df_Rt_value_pop)
```

```{r optimising_boe_parameters}
# optimising immunity_boe function

# ---- |-grid search  ----

f = function(...) compute_Rt_without_immunity(...,
                                              df_Rt_value_pop,
                                              for_optim = T,
                                              p=p)
grid_i = 1
df_grid = list()
7^4
for (NE_and_dark_factor_i in seq(from=45,to=60,length.out=15) ) {
  for ( wane_imm_fract_weekly_1_log_i in seq(from=-3,to=-0.2,length.out=15) ) {
    for ( Imm_2_partial_i in seq(from=0.5,to=0.7,length.out=15) ) {
      for ( wane_imm_fract_weekly_2_log_i in seq(from=-12,to=-8,length.out=8) ) {
        
        #######################################
        target_i = f( par=c(NE_and_dark_factor_i,
                                                      wane_imm_fract_weekly_1_log_i,
                                                      Imm_2_partial_i,
                                                      wane_imm_fract_weekly_2_log_i) 
                                                )
        df_grid[[grid_i]] = tibble(NE_and_dark_factor=NE_and_dark_factor_i,
                                   wane_imm_fract_weekly_1_log=wane_imm_fract_weekly_1_log_i,
                                   Imm_2_partial = Imm_2_partial_i,
                                   wane_imm_fract_weekly_2_log=wane_imm_fract_weekly_2_log_i,
                                   target=target_i)
        grid_i = grid_i + 1
        #########################################
      }
      
    }
  }
}
df_grid_all = bind_rows(df_grid)
df_grid_all %>% ggplot(aes(x=NE_and_dark_factor,y=target)) + geom_point() + scale_y_log10()
df_grid_all %>% ggplot(aes(x=wane_imm_fract_weekly_1_log,y=target)) + geom_point()+ scale_y_log10()
df_grid_all %>% ggplot(aes(x=Imm_2_partial,y=target)) + geom_point()+ scale_y_log10()
df_grid_all %>% ggplot(aes(x=wane_imm_fract_weekly_2_log,y=target)) + geom_point()+ scale_y_log10()

df_min = df_grid_all %>% filter(target==min(target)) 
df_Rt_without_imm = compute_Rt_without_immunity(par=c(df_min$NE_and_dark_factor,
                                                      df_min$wane_imm_fract_weekly_1_log,
                                                      df_min$Imm_2_partial,
                                                      df_min$wane_imm_fract_weekly_2_log) , 
                                                df_Rt_value_pop , 
                                                for_optim=F,
                                                p=p_loop)

c(df_min$NE_and_dark_factor,
  df_min$wane_imm_fract_weekly_1_log,
  df_min$Imm_2_partial,
  df_min$wane_imm_fract_weekly_2_log) %>% dput() %>% round(2) #  50.36  -0.40   0.67 -10.86



```

```{r}
# look
date_start = "2022-11-12"

tibble(
  date_range=c("ii_range","transmission_focus","Rt_zero"),
  date_lower=c(range_df_ii[1],transmission_focus_dates[1],Rt_zero_1),
  date_upper=c(range_df_ii[2],transmission_focus_dates[2],Rt_zero_2)) %>% 
  ggplot(aes(y=date_range,yend=date_range)) +
  geom_segment(aes(x=date_lower,xend=date_upper))

df_Rt_without_imm %>% 
  filter(date>date_start) %>% 
  ggplot(aes(x=date)) + 
  geom_line(aes(y=s3),col="orange") + 
  geom_line(aes(y=1-1/s2),col="blue") + 
  facet_wrap(~location,scales="free_y") + 
  theme(axis.ticks.y=element_blank()) + labs(subtitle = "Immunity from past (orange) + new immunity (blue)")


df_Rt_without_imm %>% 
  filter(date>date_start) %>% 
  ggplot(aes(x=date)) + 
  geom_line(aes(y=Rt),col="black") + 
  geom_hline(yintercept = 1,linetype="dashed") + facet_wrap(~location,scales="free_y") + 
  theme(axis.ticks.y=element_blank()) + labs(subtitle = "Rt empirical, estimated from infection intensity indicator")
df_Rt_without_imm %>% 
  filter(date>date_start) %>% 
  ggplot(aes(x=date)) + 
  geom_line(aes(y=Rt),col="black") + 
  geom_line(aes(y=s2*Rt),col="blue") + 
  geom_hline(yintercept = 1,linetype="dashed") + facet_wrap(~location,scales="free_y") + 
  theme(axis.ticks.y=element_blank()) + labs(subtitle = "Rt empirical (black) and accounting for new immunity only (blue)")

df_Rt_without_imm %>% 
  filter(date>date_start) %>% 
  ggplot(aes(x=date)) + 
  geom_line(aes(y=Rt_without_immunity),col="purple") + 
  geom_hline(yintercept = 1,linetype="dashed") + facet_wrap(~location,scales="free_y") + 
  theme(axis.ticks.y=element_blank()) + labs(subtitle = "Rt as if no past or new immunity")



```

### Fit augmented Rt

```{r echo=TRUE, fig.height=10, fig.width=10, message=TRUE}

# ---- |-Prepare df for stan ----
df_stan = df_Rt_without_imm %>% ungroup() %>% 
  filter(!is.na(Rt)) %>% 
  mutate(season_t=as.numeric(date-ymd("2022-12-31")) ) %>% 
  mutate(id=c(1:n()))

# ---- |-Create stan list ----
stan_list = list(
  # model input related to data-points, country structure, etc
  n = nrow(df_stan),
  n_location = n_distinct(df_stan$location),
  location_id = fct_inorder(df_stan$location) %>% as.numeric(),
  location_id_raw = fct_inorder(df_stan$location) %>% levels() %>% enframe,
  # model input related to seasonality and outcome
  season_t = df_stan$season_t,
  y = df_stan$Rt_without_immunity,
  # model input related to prior information
  prior = 2
)

checking_stan_list(stan_list)

# ---- |-Create stan list ----
fit_season=rstan::stan(
  file="../../stan/season00_start.stan",
  chains=8 ,thin=8,iter=400,# 1200,
  seed=12, cores = getOption("mc.cores", 1L),
  control=list(
    #adapt_delta=0.9,
    # max_treedepth=14
  ),
  data=stan_list
) # 
save(fit_season,stan_list,file="../../../Big data/fit_season.Rdata")
load(file="../../../Big data/fit_season.Rdata")
fit = fit_season

library(rethinking)
precis(fit,pars=c("season_fold","season_amp"),depth=2) # 40-50% more beta in winter compared to summer
precis(fit,pars=c("season_fold_log_mu","season_fold_log_sd","y_base","sigma"),depth=2)

# ---- |-Predictive checks ----
df_pp = fit %>% gather_draws(mu[id]) %>% filter(.draw%in%c(1:50)) %>% 
  left_join(df_stan , by="id")

df_stan %>% 
  ggplot(aes(x=date,y=Rt_without_immunity)) + 
  geom_line() + geom_hline(yintercept = 1) + 
  geom_line(data=df_pp,aes(x=date,y=.value,group=.draw),alpha=0.2,col="blue" ) +
  facet_wrap(~location,scales="free_y") + 
  theme(axis.text.y = element_blank(),axis.ticks.y=element_blank())
```
